---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)

# load raster
library(raster)
```

[![Build Status](https://travis-ci.org/khufkens/foto.svg)](https://travis-ci.org/khufkens/foto) [![codecov](https://codecov.io/gh/khufkens/foto/branch/master/graph/badge.svg)](https://codecov.io/gh/khufkens/foto)

# FOTO

The FOTO (Fourier Transform Textural Ordination) method uses a principal component analysis (PCA) on radially averaged 2D fourier spectra to characterize (greyscale) image texture. The FOTO method was used in [Couteron et al. 2005](http://onlinelibrary.wiley.com/doi/10.1111/j.1365-2664.2005.01097.x/abstract;jsessionid=359DD0662C272A59AF94FAEF3F213156.f02t04), [Proisy et al. 2007](http://linkinghub.elsevier.com/retrieve/pii/S0034425707000430), [Barbier et al. 2010](http://doi.wiley.com/10.1111/j.1466-8238.2009.00493.x) and [Ploton et al. 2012](http://www.esajournals.org/doi/abs/10.1890/11-1606.1). Although the techiques as presented in these papers is applied on a canopy level, the principle works on images of all types.

## Installation

### stable release

No stable release on CRAN, working on it.

### development release

To install the development releases of the package run the following commands:

```{r eval = FALSE}
if(!require(devtools)){install.package("devtools")}
devtools::install_github("khufkens/foto")
library("foto")
```

Vignettes are not rendered by default, if you want to include additional documentation please use:

```{r eval = FALSE}
if(!require(devtools)){install.package("devtools")}
devtools::install_github("khufkens/foto", build_vignettes = TRUE)
library("foto")
```

## Use

To classify image texture using the FOTO algorithm use the `foto()` function.

The raw rspectra are stored in a table called, output. The normalized values are located in noutput. The results of the principal component ordination is stored in the pcfit variable. A final classification image is stored in a rasterBrick called RGB.

<details><summary>detailed parameter description (click to expand)</summary>
<p>

Parameter     | Description                      
------------- | ------------------------------ 	
x	      | a raster layer (stack or brick)
window_size	      | a window size in pixels
plot           | plot output (TRUE / FALSE)
normalize           | normalize the radial spectrum (TRUE / FALSE)
method | "zones" or "mw" (i.e. moving window)
</p>
</details>

### Zones

The original implementation used discrete zones (blocks of x pixels wide, window_size parameter) to classify an image. This original implementation is the default, and the least computationally intensive, as it effectively reduces to resolution of the orignal data. In short, data is aggregated at the size of the specified window. 

The `foto()` routine returns a nested list with the source data (aggregated zones used and fourier components used in the PCA analysis) and a final colour image consisting of the three major principal components for every pixel.

An example analysis is run below. In the resulting image pixels with a similar colour have a similar texture. The analysis is run a historical image of plantations near Yangambi, DR Congo. The regular pattern of planted trees is picked up readily by the algorithm.

```{r figure_1, eval = TRUE, message = FALSE, warning = FALSE, fig.path = "docs/"}
# load the library
library(foto)

# load demo data
r <- raster::raster(sprintf("%s/extdata/yangambi.png", path.package("foto")))

# classify pixels using zones (discrete steps)
output <- foto(r,
     plot = TRUE,
     window_size = 25,
     method = "zones")

# print data structure
print(names(output))
```


### Moving window

To maintain the resolution of the original image a moving window approach can be used (method = "mw"). This approach overlays a window of size x (window_size parameter) on every pixel in the image and applies the FOTO methodology. This obviously represents a considerable computational burden and should be used with caution. An example is given below for a smaller subsection of the processed image above. The output format of the moving window analysis is consistent with that of the zoned approach.

```{r figure_2, eval = TRUE, fig.path = "docs/"}
# crop the image for speed
r <- crop(r, extent(1,100,1,100))

# crop the image
output <- foto(r,
     plot = TRUE,
     window_size = 25,
     method = "mw")
```

## Partitioned normalization

Partiotioned normalization as described in [Barbier et al. 2010](http://doi.wiley.com/10.1111/j.1466-8238.2009.00493.x) is not provided but easily accomplished once all images are processed. I refer to this paper for the appropriate routines.

## References




## Acknowledgements

This package is supported through the Belgian Science Policy office COBECORE project (BELSPO; grant BR/175/A3/COBECORE).