---
title: "foto background"
author: "Koen Hufkens"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{foto background}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# load the library
library(foto)
library(terra)
library(ggplot2)
library(dplyr)
library(tidyr)
```

The FOTO (Fourier Transform Textural Ordination) method uses a principal component analysis (PCA) on radially averaged 2D fourier spectra to characterize (greyscale) image texture (of canopies). Here, I'll explain the underlying algorithm, with a worked example. This should give you a better insight in how the algorithm works and how some of the parameters influence your analysis.

```{r echo = FALSE, warning=FALSE, fig.height=7, fig.width=6}
# load demo data
r <- terra::rast(sprintf("%s/extdata/yangambi.png", path.package("foto")))
terra::plot(
        r,
        col = grDevices::gray(0:100 / 100),
        legend = FALSE,
        axes = FALSE
      )
rect(50, 250, 150, 350, border = "yellow", lwd = 2)
rect(50, 550, 150, 650, border = "yellow", lwd = 2)
rect(550, 950, 650, 1050, border = "yellow", lwd = 2)
```


Note that the three boxes have very different textures.

```{r echo = FALSE, fig.height=4, fig.width=12}
im_1 <- im_2 <- im_3 <- r

im_1 <- crop(im_1, c( 50, 151, 250, 351))
im_2 <- crop(im_2, c( 50, 151, 550, 651))
im_3 <- crop(im_3, c( 550, 651, 950, 1051))


par(mfrow = c(1,3))


plot(
  im_1,
  col = grDevices::gray(0:100 / 100),
  legend = FALSE,
  axes = FALSE
)

plot(
  im_2,
  col = grDevices::gray(0:100 / 100),
  legend = FALSE,
  axes = FALSE
)

plot(
  im_3,
  col = grDevices::gray(0:100 / 100),
  legend = FALSE,
  axes = FALSE
)

im_1 <- matrix(im_1, nrow = nrow(im_1), ncol = ncol(im_1), byrow = TRUE)
im_2 <- matrix(im_2, nrow = nrow(im_2), ncol = ncol(im_2), byrow = TRUE)
im_3 <- matrix(im_3, nrow = nrow(im_3), ncol = ncol(im_3), byrow = TRUE)
```

The next step in the algorithm is to transform the data using a Fourier decomposition.

```{r}
# fft transform
fftim_1 <- Mod(stats::fft(im_1))^2
fftim_2 <- Mod(stats::fft(im_2))^2
fftim_3 <- Mod(stats::fft(im_3))^2
```

The data now need to be radially averaged, so we calculate the distance relative to the center of the image.

```{r}
# calculate distance from the center of the image
offset <- ceiling(dim(im_1)[1] / 2)

# define distance matrix
# r is rounded to an integer by zonal
distance_mask <- sqrt((col(im_1) - offset)^2 + (row(im_1) - offset)^2)
```

We first back convert all data to `terra` objects (rasters), to make the raster math easier.

```{r}
# suppress warning on missing extent (which
# is normal)
fftim_1 <- terra::rast(fftim_1)
fftim_2 <- terra::rast(fftim_2)
fftim_3 <- terra::rast(fftim_3)
distance_mask <- terra::rast(distance_mask)
```

Below you see the concentric mask and the 2D FFT frequency plot. Note, unlike most conventional plots the data is not centered on 0.

```{r  fig.height = 6, fig.width=12}
par(mfrow=c(1,2))
plot(log(fftim_1))
plot(distance_mask)

```

For all FFT spectra the data will be radially averaged, removing any directional components from the data. The `terra` zonal() function is used to do this efficiently.

```{r}
rspec_1 <- rev(
  terra::zonal(
    fftim_1,
    distance_mask,
    fun = "mean",
    na.rm = TRUE
  )[, 2]
)
```

```{r echo = FALSE}
rspec_2 <- rev(
  terra::zonal(
    fftim_2,
    distance_mask,
    fun = "mean",
    na.rm = TRUE
  )[, 2]
)

rspec_3 <- rev(
  terra::zonal(
    fftim_3,
    distance_mask,
    fun = "mean",
    na.rm = TRUE
  )[, 2]
)

df <- data.frame(
  idx = 1:length(rspec_1),
  box_1 = rspec_1,
  box_2 = rspec_2,
  box_3 = rspec_3
) |>
  pivot_longer(
    cols = starts_with("box"),
    values_to = "spec",
    names_to = "box"
  )
```

Plotting the averaged spectra shows subtle differences between the different textures.

```{r fig.height=4, fig.width=12, echo = FALSE}
p  <- ggplot(df) +
  geom_line(
    aes(
      idx,
      log(spec),
      group = box,
      colour = box
    )
  )

print(p)

```